---
title: "Inspect and prepare a Shiny app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspect and prepare a Shiny app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  width = 72
)
```

`rpackit` separates packaging into a few explicit checks. This makes it
possible to understand an application before copying a runtime or installing
packages:

1. `doctor()` checks the local build environment.
2. `check_app()` identifies the application layout and suitable targets.
3. `plan_dependencies()` explains which R packages a bundle needs and why.
4. `prepare_desktop()` creates portable desktop resources atomically.
5. `validate_desktop_bundle()` verifies those resources without running the
   application.
6. `start_desktop_app()` and `stop_desktop_app()` provide a managed,
   authenticated development lifecycle.

## Create a small application

The example below creates an ordinary single-file Shiny application in the
R session's temporary directory.

```{r example-app}
library(rpackit)

app <- tempfile("hello-rpackit-")
dir.create(app)
writeLines(
  c(
    "library(shiny)",
    "shinyApp(",
    "  fluidPage(h1('Hello from rpackit')),",
    "  function(input, output, session) {}",
    ")"
  ),
  file.path(app, "app.R")
)
```

## Inspect before building

```{r inspect}
inspection <- check_app(app, quiet = TRUE)
inspection$app_type
inspection$targets
inspection$findings$system_calls
```

The target matrix is evidence, not a promise that every dependency works in
every target. In particular, a `maybe` result for static web means that the
application still needs a real shinylive/webR export test.

External commands need special attention in a portable build. When source code
directly calls `system()`, `system2()`, or `shell()`,
`inspection$findings$system_calls` identifies the call, file, and line.
Comments and quoted string contents are ignored because rpackit inspects
parsed R syntax rather than searching raw text. Calls inside quoted language
expressions are reported conservatively because static inspection cannot know
whether the application evaluates them later.

## Review the dependency plan

```{r dependencies}
plan <- plan_dependencies(app)
plan$dependencies
plan$diagnostics
```

The planner combines parsed source calls, `DESCRIPTION`, and `renv.lock`
without evaluating application code. Start with `diagnostics`: a dynamic
package name such as `library(pkg, character.only = TRUE)` cannot be resolved
statically and should be made explicit before packaging.

When present, `renv.lock` supplies exact package versions and sources.
`DESCRIPTION` supplies direct dependency roles and constraints. Parsed source
calls catch packages that were used but not declared. The complete evidence is
available in `plan$references`.

## Check the build machine

```{r doctor}
environment <- doctor(quiet = TRUE)
environment$platform
environment$capabilities
```

Missing tools do not prevent read-only inspection. They only affect the build
targets that require them. For example, a Windows Tauri build needs Node,
Cargo, the Tauri CLI, a JavaScript package manager, the native toolchain, and
WebView2.

## Prepare portable desktop resources

On a platform covered by the verified portable-R registry, rpackit can resolve
and cache the runtime automatically:

```{r prepare, eval = FALSE}
bundle <- prepare_desktop(
  app,
  runtime_dir = NULL
)

validation <- validate_desktop_bundle(
  bundle$path,
  verify_runtime = TRUE
)
validation
```

Set `offline = TRUE` in `prepare_desktop()` when a matching runtime is already
cached and the build must not read the registry or artifact. Alternatively,
pass an extracted portable R home as `runtime_dir`.

Preparation never overwrites an existing output directory. It assembles a
sibling staging directory, validates it, and only then publishes the completed
resource tree. If preparation fails, fix the reported input or dependency
problem and choose a new or removed output directory before retrying.

## Run and stop the prepared app

The R-level lifecycle is intended for trusted development and native
integration testing:

```{r lifecycle, eval = FALSE}
process <- start_desktop_app(bundle$path)
on.exit(stop_desktop_app(process, quiet = TRUE), add = TRUE)

desktop_app_status(process)
launch <- desktop_app_launch_config(process)

# Hand `launch` directly to a trusted native consumer.
# Never print, log, serialize, or persist launch$headers.

stop_desktop_app(process)
```

The endpoint requires a request header that a stock browser cannot add to
top-level navigation and WebSocket requests. Do not paste the token-free URL
from `desktop_app_status()` into a normal browser and do not put the secret in
the URL. Always stop the process, including on errors; the `on.exit()` pattern
above supplies that cleanup path.

`prepare_desktop()` currently creates the portable resource contract. It does
not yet generate a supported Tauri installer. The
[rpackit roadmap](https://github.com/rpackit/roadmap) records the native
transport and release gates separately from this R workflow.
